---
title: "ARMSPACKAGE: Spatial Copy Number Analysis Workflow"
author: "ARMSPACKAGE Development Team"
date: "`r Sys.Date()`"
output: html_notebook
vignette: >
  %\VignetteIndexEntry{ARMSPACKAGE: Spatial Copy Number Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

ARMS..

## Key Features

- Load and align ASCAT.sc copy number profiles to a common genomic grid
- Hierarchical clustering with dynamic tree cutting to identify clonal populations
- UMAP visualization of sample relationships
- Spatial mapping integration with GeoJSON data
- Pseudobulk analysis for improved signal-to-noise ratio
- Quality metrics and diagnostic plots
- Clone refinement by correlation
- Bulk profile simulation

# Installation

```{r install, eval=FALSE}
# From Bioconductor (after acceptance)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ARMSPACKAGE")

# From GitHub (development version)
devtools::install_github("yourusername/ARMSPACKAGE")
```

# Load Package

```{r load}
library(ARMS)
```

# Workflow Overview

The typical workflow consists of these steps:

1. **Run ASCAT.sc** on BAM files (using ASCAT.sc directly)
2. **Load profiles** into R matrices
3. **Diagnostic plots** for quality control
4. **Hierarchical clustering** to identify clones
5. **Visualization** with heatmaps and UMAP
6. **Spatial analysis** (optional)
7. **Pseudobulk analysis** (optional)
8. **Clone refinement** (optional)
9. **Bulk profile analysis** (optional)

# Complete Workflow

## Step 1: Run ASCAT.sc on BAM Files

The first step is running ASCAT.sc to generate copy number profiles from BAM files.

```{r ascat, eval=FALSE}
# Get example BAM files from package
bam_dir <- system.file("extdata/dupmarked_bam/", package = "ARMS")
bam_files <- list.files(bam_dir, pattern = "\\.bam$", full.names = TRUE)

# Or use your own BAM files
# bam_files <- list.files("path/to/bams/", pattern = "\\.bam$", full.names = TRUE)

# Run ASCAT.sc on BAM files
# This step generates LogR and copy number profiles for each cell/spot
# Note: This can take 5-15 minutes per sample
# make sure ascat output dir exists, otherwise create in extdata folder
ascat_output_dir = paste0("inst/extdata/", "ascat_output/")
if (!dir.exists(ascat_output_dir)) {
  dir.create(ascat_output_dir, recursive = TRUE)
}
project_name = "example"



ascat_result <- ASCAT.sc::run_sc_sequencing(
  tumour_bams = bam_files,
  allchr = paste0("chr", c(1:22, "X")),
  sex = rep("male", length(bam_files)),
  chrstring_bam = "chr",
  binsize = 30000,
  purs = seq(0.01, 1, 0.01),
  ploidies = seq(1.7, 5, 0.01),
  outdir = ascat_output_dir,
  projectname = project_name,
  build = "hg38"
)

# Profiles are saved to: ascat_output/profiles/
```

## Step 2: Load ASCAT Profiles

Load the ASCAT.sc profiles into R:

```{r profiles, eval=FALSE}
# Load profiles generated by ASCAT.sc
profiles <- load_ascat_profiles(
  profile_dir = ascat_output_dir,
  pattern = "*refitted.ASCAT.scprofile.txt",
  resolution = 30000
)

# Inspect what was loaded
names(profiles)
# [1] "copy_number_matrix" "logr_matrix" "grid" "sample_names"

dim(profiles$logr_matrix)
# [1]  150 3247   (150 samples, 3247 genomic bins)
```

For this vignette, we'll load ASCAT.sc output from the example data:

```{r example_data}
# Path to example ASCAT.sc output
#ascat_output_dir <- system.file("extdata/ascat_output", package = "ARMS")

# Load profiles from example data
profiles <- load_ascat_profiles(
  profile_dir = ascat_output_dir,
  pattern = "*example_refitted.ASCAT.scprofile.txt",
  resolution = 30000
)

# Extract logR matrix and genomic grid for downstream analysis
logr_matrix <- profiles$logr_matrix
genomic_grid <- profiles$grid

dim(logr_matrix)
head(genomic_grid)
```

## Step 3: Diagnostic Plots (QC)

Before clustering, examine data quality:

```{r diagnostics, eval=FALSE}
# Ploidy distribution across samples
plot_ploidy_distribution(
  copy_number_matrix = profiles$copy_number_matrix,
  output_file = "ploidy_distribution.pdf"
)

# LogR variance per sample (QC metric)
plot_logr_variance(
  logr_matrix = logr_matrix,
  output_file = "logr_variance.pdf"
)

# Chromosomal instability index
plot_chromosomal_instability(
  copy_number_matrix = profiles$copy_number_matrix,
  grid = genomic_grid,
  output_file = "chromosomal_instability.pdf"
)
```

## Step 4: Hierarchical Clustering

Identify clonal populations using hierarchical clustering with dynamic tree cutting:

```{r clustering}
# Perform hierarchical clustering
clustering <- hierarchical_clustering(
  logr_matrix = logr_matrix,
  method = "ward.D2",
  min_cluster_size = 3
)

# View cluster assignments
table(clustering$clusters)

# Clustering automatically determines optimal number of clusters
print(paste("Found", max(clustering$clusters), "clusters"))
```

**Key parameters:**
- `method`: Linkage method ("ward.D2" recommended for compact clusters)
- `min_cluster_size`: Minimum cells required per cluster
- `use_dynamic`: Use dynamic tree cutting with recursive splitting (default: TRUE)

## Step 5: Visualization - Heatmaps and UMAP

### Heatmap

Generate a LogR heatmap showing copy number patterns:

```{r heatmap, eval=FALSE}
# Create heatmap with cluster annotations
plot_logr_heatmap(
  logr_matrix = logr_matrix,
  grid = genomic_grid,
  clusters = clustering$clusters,
  output_file = "logr_heatmap.pdf",
  width = 12,
  height = 8
)
```

```{r heatmap_small, echo=FALSE, fig.cap="LogR heatmap showing clonal populations"}
# Show smaller version for vignette
plot_logr_heatmap(
  logr_matrix = logr_matrix[1:50, ],
  grid = genomic_grid,
  clusters = clustering$clusters[1:50],
  output_file = "logr_heatmap_subset.pdf"
)
```

### UMAP

Visualize cells in 2D space:

```{r umap}
# Compute UMAP embedding
umap_coords <- compute_umap(
  logr_matrix = logr_matrix,
  n_neighbors = 15,
  min_dist = 0.1
)

# Plot with cluster colors
plot_umap(
  umap_coords = umap_coords,
  clusters = clustering$clusters,
  output_file = "umap.pdf"
)
```

## Step 6: Spatial Visualization (Optional)

If you have GeoJSON spatial coordinates:

```{r spatial, eval=FALSE}
# Load spatial tile data
spatial_tiles <- load_geojson_tiles(
  geojson_file = "tissue_positions.geojson"
)

# Match samples to tiles
spatial_matches <- match_samples_to_tiles(
  sample_ids = rownames(logr_matrix),
  tile_names = spatial_tiles$tile_name
)

# Plot spatial distribution
plot_spatial_clusters(
  tiles = spatial_tiles,
  matches = spatial_matches,
  clusters = clustering$clusters,
  output_file = "spatial_clusters.pdf"
)

# Calculate spatial autocorrelation (Moran's I)
morans_i <- calculate_morans_i(
  tiles = spatial_tiles,
  matches = spatial_matches,
  clusters = clustering$clusters
)

# Plot spatial coherence
plot_spatial_coherence(morans_i, output_file = "spatial_coherence.pdf")
```

# Optional Steps

These mirror the optional steps in the Snakemake workflow (controlled by config flags).

## Step 7: Pseudobulk Analysis (Optional)

Create pseudobulk profiles for better signal-to-noise ratio by merging cells within
each cluster:

```{r pseudobulk, eval=FALSE}
# Fast pseudobulk from bin counts (no BAM merging required!)
pseudobulk_tracks <- create_pseudobulk_from_bins(
  ascat_rda_file = file.path(ascat_output_dir, paste0(project_name, ".Rda")),
  clusters = clustering$clusters,
  sample_names = profiles$sample_names,
  output_dir = "pseudobulk_output",
  min_samples = 3
)

# Run ASCAT on pseudobulk tracks
pseudobulk_ascat <- run_ascat_on_pseudobulk_tracks(
  pseudobulk_tracks = pseudobulk_tracks,
  ascat_rda_file = file.path(ascat_output_dir, paste0(project_name, ".Rda")),
  output_dir = "pseudobulk_ascat"
)

# Compare single-cell vs pseudobulk quality
quality_comparison <- compare_sc_vs_pseudobulk(
  sc_profiles = profiles,
  pb_profiles = pseudobulk_ascat,
  output_file = "quality_comparison.pdf"
)

# Visualize pseudobulk alongside single-cell
plot_sc_pb_comparison_heatmap(
  sc_logr = logr_matrix,
  pb_logr = pseudobulk_ascat$logr_matrix,
  grid = genomic_grid,
  clusters = clustering$clusters,
  output_file = "sc_pb_comparison.pdf"
)
```

## Step 8: Clone Refinement (Optional)

Merge highly correlated clusters based on correlation threshold:

```{r refine, eval=FALSE}
# Refine by merging similar clusters
refined_clusters <- refine_clusters_by_correlation(
  logr_matrix = logr_matrix,
  clusters = clustering$clusters,
  correlation_threshold = 0.9
)

table(refined_clusters)

# Visualize cluster correlation
plot_cluster_correlation(
  logr_matrix = logr_matrix,
  clusters = clustering$clusters,
  output_file = "cluster_correlation.pdf"
)

# Re-plot heatmap with refined clusters
plot_logr_heatmap(
  logr_matrix = logr_matrix,
  grid = genomic_grid,
  clusters = refined_clusters,
  output_file = "logr_heatmap_refined.pdf"
)
```

## Step 9: Bulk Profile Analysis (Optional)

Simulate bulk sequencing by merging all cells:

```{r bulk, eval=FALSE}
# Create bulk profile
bulk_track <- create_bulk_from_bins(
  ascat_rda_file = file.path(ascat_output_dir, paste0(project_name, ".Rda")),
  output_dir = "bulk_output"
)

# Run ASCAT on bulk
bulk_ascat <- run_ascat_on_bulk_track(
  bulk_track = bulk_track,
  ascat_rda_file = file.path(ascat_output_dir, paste0(project_name, ".Rda")),
  output_dir = "bulk_ascat"
)

# Visualize bulk profile
plot_logr_heatmap(
  logr_matrix = bulk_ascat$logr_matrix,
  grid = genomic_grid,
  clusters = rep(1, nrow(bulk_ascat$logr_matrix)),
  output_file = "bulk_logr_heatmap.pdf"
)
```

# Complete Example Workflow

Here's a complete end-to-end example from loading profiles through all analyses:

```{r complete_example, eval=FALSE}
library(ARMS)

# Load ASCAT.sc profiles
profiles <- load_ascat_profiles(
  profile_dir = "ascat_output/profiles",
  pattern = "*refitted.ASCAT.scprofile.txt",
  resolution = 30000
)

# Quality control plots
plot_ploidy_distribution(profiles$copy_number_matrix, "ploidy.pdf")
plot_logr_variance(profiles$logr_matrix, "variance.pdf")
plot_chromosomal_instability(profiles$copy_number_matrix, profiles$grid, "instability.pdf")

# Hierarchical clustering
clustering <- hierarchical_clustering(
  logr_matrix = profiles$logr_matrix,
  method = "ward.D2",
  min_cluster_size = 3
)

# Visualize
plot_logr_heatmap(profiles$logr_matrix, profiles$grid, clustering$clusters, "heatmap.pdf")
umap_coords <- compute_umap(profiles$logr_matrix)
plot_umap(umap_coords, clustering$clusters, "umap.pdf")

# Spatial analysis (if spatial data available)
tiles <- load_geojson_tiles("tissue_positions.geojson")
matches <- match_samples_to_tiles(rownames(profiles$logr_matrix), tiles$tile_name)
plot_spatial_clusters(tiles, matches, clustering$clusters, "spatial.pdf")
morans_i <- calculate_morans_i(tiles, matches, clustering$clusters)
plot_spatial_coherence(morans_i, "spatial_coherence.pdf")

# Pseudobulk analysis
pseudobulk_tracks <- create_pseudobulk_from_bins(
  ascat_rda_file = "ascat_output/example.Rda",
  clusters = clustering$clusters,
  sample_names = profiles$sample_names,
  output_dir = "pseudobulk_output",
  min_samples = 3
)

pseudobulk_ascat <- run_ascat_on_pseudobulk_tracks(
  pseudobulk_tracks = pseudobulk_tracks,
  ascat_rda_file = "ascat_output/example.Rda",
  output_dir = "pseudobulk_ascat"
)

plot_sc_pb_comparison_heatmap(
  profiles$logr_matrix,
  pseudobulk_ascat$logr_matrix,
  profiles$grid,
  clustering$clusters,
  "sc_pb_comparison.pdf"
)

# Clone refinement
refined_clusters <- refine_clusters_by_correlation(
  logr_matrix = profiles$logr_matrix,
  clusters = clustering$clusters,
  correlation_threshold = 0.9
)

plot_logr_heatmap(profiles$logr_matrix, profiles$grid, refined_clusters, "heatmap_refined.pdf")
plot_cluster_correlation(profiles$logr_matrix, clustering$clusters, "cluster_correlation.pdf")

# Bulk profile simulation
bulk_track <- create_bulk_from_bins(
  ascat_rda_file = "ascat_output/example.Rda",
  output_dir = "bulk_output"
)

bulk_ascat <- run_ascat_on_bulk_track(
  bulk_track = bulk_track,
  ascat_rda_file = "ascat_output/example.Rda",
  output_dir = "bulk_ascat"
)
```

# Tips and Best Practices

## Choosing Parameters

**Binsize:**
- **30kb**: Optimal for spatial transcriptomics (10x Visium)
- **100kb**: Standard for low-coverage single-cell
- **1Mb**: For very sparse data or whole-chromosome analysis

**Clustering:**
- Start with `min_cluster_size = 3-5` to avoid spurious clusters
- Use `method = "ward.D2"` for tight, spherical clusters
- Set `use_dynamic = TRUE` for automatic cluster detection (default)

**Cluster refinement:**
- Use `correlation_threshold = 0.9-0.95` to merge similar clones
- Lower threshold = more aggressive merging

## Quality Control

- Remove samples with abnormally high LogR variance (technical noise)
- Check ploidy distribution for unexpected values (contamination, normal cells)
- Verify spatial coherence - clones should be spatially contiguous

## Performance

- Use `cores` parameter for parallel processing
- Pseudobulk analysis is faster and more robust than single-cell
- For large datasets (>500 cells), consider downsampling for initial QC

## Troubleshooting

**Too many small clusters:**
- Increase `min_cluster_size`
- Use cluster refinement with `refine_clusters_by_correlation()`

**Poor spatial coherence:**
- May indicate true spatial heterogeneity
- Check for batch effects or technical artifacts
- Consider using pseudobulk profiles

**Memory issues:**
- Reduce genomic resolution (increase binsize)
- Process chromosomes separately
- Use Snakemake workflow with checkpointing

# Session Info

```{r sessioninfo}
sessionInfo()
```
