---
title: "ARMSPACKAGE: Spatial Copy Number Analysis Workflow"
author: "ARMSPACKAGE Development Team"
date: "`r Sys.Date()`"
output: html_notebook
vignette: >
  %\VignetteIndexEntry{ARMSPACKAGE: Spatial Copy Number Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

ARMS..

## Key Features

- Load and align ASCAT.sc copy number profiles to a common genomic grid
- Hierarchical clustering with dynamic tree cutting to identify clonal populations
- UMAP visualization of sample relationships
- Spatial mapping integration with GeoJSON data
- Pseudobulk analysis for improved signal-to-noise ratio
- Quality metrics and diagnostic plots
- Clone refinement by correlation
- Bulk profile simulation

# Installation

```{r install, eval=FALSE}
# From Bioconductor (after acceptance)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ARMSPACKAGE")

# From GitHub (development version)
devtools::install_github("yourusername/ARMSPACKAGE")
```

# Load Package

```{r load}
library(ARMS)
```

# Workflow Overview

The typical workflow consists of these steps:

1. **Run ASCAT.sc** on BAM files (using ASCAT.sc directly)
2. **Load profiles** into R matrices
3. **Diagnostic plots** for quality control
4. **Hierarchical clustering** to identify clones
5. **Visualization** with heatmaps and UMAP
6. **Spatial analysis** (optional)
7. **Pseudobulk analysis** (optional)
8. **Clone refinement** (optional)
9. **Bulk profile analysis** (optional)

# Complete Workflow

## Step 1: Run ASCAT.sc on BAM Files

The first step is running ASCAT.sc to generate copy number profiles from BAM files.

```{r ascat, eval=FALSE}
# Get example BAM files from package
bam_dir <- system.file("extdata/dupmarked_bam/", package = "ARMS")
bam_files <- list.files(bam_dir, pattern = "\\.bam$", full.names = TRUE)

# Or use your own BAM files
# bam_files <- list.files("path/to/bams/", pattern = "\\.bam$", full.names = TRUE)

# Run ASCAT.sc on BAM files
# This step generates LogR and copy number profiles for each cell/spot
# Note: This can take 5-15 minutes per sample
# make sure ascat output dir exists, otherwise create in extdata folder
ascat_output_dir = paste0("inst/extdata/", "ascat_output/")
if (!dir.exists(ascat_output_dir)) {
  dir.create(ascat_output_dir, recursive = TRUE)
}
project_name = "example"



ascat_result <- ASCAT.sc::run_sc_sequencing(
  tumour_bams = bam_files,
  allchr = paste0("chr", c(1:22, "X")),
  sex = rep("male", length(bam_files)),
  chrstring_bam = "chr",
  binsize = 30000,
  purs = seq(0.01, 1, 0.01),
  ploidies = seq(1.7, 5, 0.01),
  outdir = ascat_output_dir,
  projectname = project_name,
  build = "hg38"
)

# Profiles are saved to: ascat_output/profiles/
```

## Step 2: Load ASCAT Profiles

Load the ASCAT.sc profiles into R:

```{r profiles, eval=FALSE}
# Load profiles generated by ASCAT.sc
profiles <- load_ascat_profiles(
  profile_dir = ascat_output_dir,
  pattern = "*refitted.ASCAT.scprofile.txt",
  resolution = 30000
)

# Inspect what was loaded
names(profiles)
# [1] "copy_number_matrix" "logr_matrix" "grid" "sample_names"

dim(profiles$logr_matrix)
# [1]  150 3247   (150 samples, 3247 genomic bins)
```

For this vignette, we'll load ASCAT.sc output from the example data:

```{r example_data}
# Path to example ASCAT.sc output
#ascat_output_dir <- system.file("extdata/ascat_output", package = "ARMS")

# Load profiles from example data
profiles <- load_ascat_profiles(
  profile_dir = ascat_output_dir,
  pattern = "*example_refitted.ASCAT.scprofile.txt",
  resolution = 30000
)

# Extract logR matrix and genomic grid for downstream analysis
logr_matrix <- profiles$logr_matrix
genomic_grid <- profiles$grid

dim(logr_matrix)
head(genomic_grid)
```

## Step 3: Diagnostic Plots (QC)

Before clustering, examine data quality:

```{r diagnostics, eval=FALSE}
# Ploidy distribution across samples
plot_ploidy_distribution(
  copy_number_matrix = profiles$copy_number_matrix,
  output_file = "ploidy_distribution.pdf"
)

# LogR variance per sample (QC metric)
plot_logr_variance(
  logr_matrix = logr_matrix,
  output_file = "logr_variance.pdf"
)

# Chromosomal instability index
plot_chromosomal_instability(
  copy_number_matrix = profiles$copy_number_matrix,
  grid = genomic_grid,
  output_file = "chromosomal_instability.pdf"
)
```

## Step 4: Clustering

Identify clonal populations using the recommended default clustering. By default,
`weighted_hierarchical_clustering()` discretizes logR values to gain/neutral/loss
categories before clustering, so samples with the same copy-number profile but
different tumor purity group together. No extra arguments are needed:

```{r clustering}
# Perform clustering (discretized by default)
clustering <- weighted_hierarchical_clustering(
  logr_matrix = logr_matrix
)

# View cluster assignments
table(clustering$clusters)

# Clustering automatically determines the number of clusters
print(paste("Found", max(clustering$clusters), "clusters"))
```

**Key parameters:**
- `discretize`: Discretize logR to gain/neutral/loss before clustering (default: TRUE)
- `method`: Linkage method ("ward.D2" recommended for compact clusters)
- `min_cluster_size`: Minimum cells required per cluster (default: 3)
- `use_dynamic`: Use dynamic tree cutting with recursive splitting (default: TRUE)

To cluster on continuous logR instead of discretized categories, use
`hierarchical_clustering(logr_matrix)` or pass `discretize = FALSE`.

## Step 5: Visualization - Heatmaps and UMAP

### Heatmap

Generate a LogR heatmap showing copy number patterns:

```{r heatmap, eval=FALSE}
# Create heatmap with cluster annotations
plot_logr_heatmap(
  logr_matrix = logr_matrix,
  grid = genomic_grid,
  clusters = clustering$clusters,
  output_file = "logr_heatmap.pdf",
  width = 12,
  height = 8
)
```

```{r heatmap_small, echo=FALSE, fig.cap="LogR heatmap showing clonal populations"}
# Show smaller version for vignette
plot_logr_heatmap(
  logr_matrix = logr_matrix[1:50, ],
  grid = genomic_grid,
  clusters = clustering$clusters[1:50],
  output_file = "logr_heatmap_subset.pdf"
)
```

### UMAP

Visualize cells in 2D space:

```{r umap}
# Compute UMAP embedding
umap_coords <- compute_umap(
  logr_matrix = logr_matrix,
  n_neighbors = 15,
  min_dist = 0.1
)

# Plot with cluster colors
plot_umap(
  umap_coords = umap_coords,
  clusters = clustering$clusters,
  output_file = "umap.pdf"
)
```

## Step 6: Spatial Visualization (Optional)

If you have GeoJSON spatial coordinates:

```{r spatial, eval=FALSE}
# Load spatial tile data
spatial_tiles <- load_geojson_tiles(
  geojson_file = "tissue_positions.geojson"
)

# Match samples to tiles
spatial_matches <- match_samples_to_tiles(
  sample_ids = rownames(logr_matrix),
  tile_names = spatial_tiles$tile_name
)

# Plot spatial distribution
plot_spatial_clusters(
  tiles = spatial_tiles,
  matches = spatial_matches,
  clusters = clustering$clusters,
  output_file = "spatial_clusters.pdf"
)

# Calculate spatial autocorrelation (Moran's I)
morans_i <- calculate_morans_i(
  tiles = spatial_tiles,
  matches = spatial_matches,
  clusters = clustering$clusters
)

# Plot spatial coherence
plot_spatial_coherence(morans_i, output_file = "spatial_coherence.pdf")
```

# Optional Steps

These mirror the optional steps in the Snakemake workflow (controlled by config flags).

## Step 7: Pseudobulk Analysis (Optional)

Create pseudobulk profiles for better signal-to-noise ratio by merging cells within
each cluster:

```{r pseudobulk, eval=FALSE}
# Fast pseudobulk from bin counts (no BAM merging required!)
pseudobulk_tracks <- create_pseudobulk_from_bins(
  ascat_rda_file = file.path(ascat_output_dir, paste0(project_name, ".Rda")),
  clusters = clustering$clusters,
  sample_names = profiles$sample_names,
  output_dir = "pseudobulk_output",
  min_samples = 3
)

# Run ASCAT on pseudobulk tracks
pseudobulk_ascat <- run_ascat_on_pseudobulk_tracks(
  pseudobulk_tracks = pseudobulk_tracks,
  ascat_rda_file = file.path(ascat_output_dir, paste0(project_name, ".Rda")),
  output_dir = "pseudobulk_ascat"
)

# Compare single-cell vs pseudobulk quality
quality_comparison <- compare_sc_vs_pseudobulk(
  sc_profiles = profiles,
  pb_profiles = pseudobulk_ascat,
  output_file = "quality_comparison.pdf"
)

# Visualize pseudobulk alongside single-cell
plot_sc_pb_comparison_heatmap(
  sc_logr = logr_matrix,
  pb_logr = pseudobulk_ascat$logr_matrix,
  grid = genomic_grid,
  clusters = clustering$clusters,
  output_file = "sc_pb_comparison.pdf"
)
```

## Step 8: Clone Refinement (Optional)

The recommended way to refine clusters is manually, by splitting clusters that contain
distinct subpopulations and merging clusters that represent the same clone. Inspect the
heatmap from Step 5, then adjust:

```{r refine, eval=FALSE}
# Split a cluster that contains distinct subpopulations
clustering_split <- split_clusters_manually(
  clustering_obj = clustering,
  clusters_to_split = 4,
  logr_matrix = logr_matrix,
  method = "weighted",
  grid = genomic_grid,
  k = 3
)

# Merge groups of clusters that represent the same clone
clustering_merged <- merge_clusters_manually(
  clustering_split,
  clusters_to_merge = list(c(2, 5), c(3, 6))
)

table(clustering_merged$clusters)

# Re-plot heatmap with refined clusters
plot_logr_heatmap(
  logr_matrix = logr_matrix,
  grid = genomic_grid,
  clusters = clustering_merged$clusters,
  output_file = "logr_heatmap_refined.pdf"
)
```

### Experimental: automatic correlation-based merging

As an experimental alternative, `refine_clusters_by_correlation()` automatically merges
clusters whose centroid profiles correlate above a threshold. This is not part of the
default workflow:

```{r refine_cor, eval=FALSE}
refined_clusters <- refine_clusters_by_correlation(
  logr_matrix = logr_matrix,
  clusters = clustering$clusters,
  cor_threshold = 0.9
)

plot_cluster_correlation(
  logr_matrix = logr_matrix,
  clusters = clustering$clusters,
  output_file = "cluster_correlation.pdf"
)
```

## Step 9: Bulk Profile Analysis (Optional)

Simulate bulk sequencing by merging all cells:

```{r bulk, eval=FALSE}
# Create bulk profile
bulk_track <- create_bulk_from_bins(
  ascat_rda_file = file.path(ascat_output_dir, paste0(project_name, ".Rda")),
  output_dir = "bulk_output"
)

# Run ASCAT on bulk
bulk_ascat <- run_ascat_on_bulk_track(
  bulk_track = bulk_track,
  ascat_rda_file = file.path(ascat_output_dir, paste0(project_name, ".Rda")),
  output_dir = "bulk_ascat"
)

# Visualize bulk profile
plot_logr_heatmap(
  logr_matrix = bulk_ascat$logr_matrix,
  grid = genomic_grid,
  clusters = rep(1, nrow(bulk_ascat$logr_matrix)),
  output_file = "bulk_logr_heatmap.pdf"
)
```

# Complete Example Workflow

Here's a complete end-to-end example from loading profiles through all analyses:

```{r complete_example, eval=FALSE}
library(ARMS)

# Load ASCAT.sc profiles
profiles <- load_ascat_profiles(
  profile_dir = "ascat_output/profiles",
  pattern = "*refitted.ASCAT.scprofile.txt",
  resolution = 30000
)

# Quality control plots
plot_ploidy_distribution(profiles$copy_number_matrix, "ploidy.pdf")
plot_logr_variance(profiles$logr_matrix, "variance.pdf")
plot_chromosomal_instability(profiles$copy_number_matrix, profiles$grid, "instability.pdf")

# Clustering (discretized by default)
clustering <- weighted_hierarchical_clustering(
  logr_matrix = profiles$logr_matrix
)

# Visualize
plot_logr_heatmap(profiles$logr_matrix, profiles$grid, clustering$clusters, "heatmap.pdf")
umap_coords <- compute_umap(profiles$logr_matrix)
plot_umap(umap_coords, clustering$clusters, "umap.pdf")

# Spatial analysis (if spatial data available)
tiles <- load_geojson_tiles("tissue_positions.geojson")
matches <- match_samples_to_tiles(rownames(profiles$logr_matrix), tiles$tile_name)
plot_spatial_clusters(tiles, matches, clustering$clusters, "spatial.pdf")
morans_i <- calculate_morans_i(tiles, matches, clustering$clusters)
plot_spatial_coherence(morans_i, "spatial_coherence.pdf")

# Pseudobulk analysis
pseudobulk_tracks <- create_pseudobulk_from_bins(
  ascat_rda_file = "ascat_output/example.Rda",
  clusters = clustering$clusters,
  sample_names = profiles$sample_names,
  output_dir = "pseudobulk_output",
  min_samples = 3
)

pseudobulk_ascat <- run_ascat_on_pseudobulk_tracks(
  pseudobulk_tracks = pseudobulk_tracks,
  ascat_rda_file = "ascat_output/example.Rda",
  output_dir = "pseudobulk_ascat"
)

plot_sc_pb_comparison_heatmap(
  profiles$logr_matrix,
  pseudobulk_ascat$logr_matrix,
  profiles$grid,
  clustering$clusters,
  "sc_pb_comparison.pdf"
)

# Clone refinement (manual, recommended)
clustering_merged <- merge_clusters_manually(
  clustering,
  clusters_to_merge = list(c(2, 5), c(3, 6))
)

plot_logr_heatmap(profiles$logr_matrix, profiles$grid, clustering_merged$clusters, "heatmap_refined.pdf")

# Bulk profile simulation
bulk_track <- create_bulk_from_bins(
  ascat_rda_file = "ascat_output/example.Rda",
  output_dir = "bulk_output"
)

bulk_ascat <- run_ascat_on_bulk_track(
  bulk_track = bulk_track,
  ascat_rda_file = "ascat_output/example.Rda",
  output_dir = "bulk_ascat"
)
```

# Tips and Best Practices

## Choosing Parameters

**Binsize:**
- **30kb**: Optimal for spatial transcriptomics (10x Visium)
- **100kb**: Standard for low-coverage single-cell
- **1Mb**: For very sparse data or whole-chromosome analysis

**Clustering:**
- `weighted_hierarchical_clustering()` is the recommended default (discretizes logR so purity differences don't split identical clones)
- Start with `min_cluster_size = 3-5` to avoid spurious clusters
- Use `method = "ward.D2"` for tight, spherical clusters
- Set `use_dynamic = TRUE` for automatic cluster detection (default)

**Cluster refinement:**
- Refine manually with `split_clusters_manually()` and `merge_clusters_manually()` after inspecting the heatmap
- Experimental alternative: `refine_clusters_by_correlation(cor_threshold = 0.9-0.95)` to auto-merge similar clones (lower threshold = more aggressive merging)

## Quality Control

- Remove samples with abnormally high LogR variance (technical noise)
- Check ploidy distribution for unexpected values (contamination, normal cells)
- Verify spatial coherence - clones should be spatially contiguous

## Performance

- Use `cores` parameter for parallel processing
- Pseudobulk analysis is faster and more robust than single-cell
- For large datasets (>500 cells), consider downsampling for initial QC

## Troubleshooting

**Too many small clusters:**
- Increase `min_cluster_size`
- Merge similar clusters with `merge_clusters_manually()` (or, experimentally, `refine_clusters_by_correlation()`)

**Poor spatial coherence:**
- May indicate true spatial heterogeneity
- Check for batch effects or technical artifacts
- Consider using pseudobulk profiles

**Memory issues:**
- Reduce genomic resolution (increase binsize)
- Process chromosomes separately
- Use Snakemake workflow with checkpointing

# Session Info

```{r sessioninfo}
sessionInfo()
```
